Left-linear rules in adhesive categories

### 2024-02-08 icalp/fusione.tex
structure
minor things - sec 1

This first section is devoted to recall the definition and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive}.

@@

In this section we will define the notion of $\mathcal{M}$-adhesivity \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,heindel2009category,lack2005adhesive} and explore some of the consequence of such a property.

@@

\rem{rimuoverei}{
  \begin{remark}
    \label{rem:diff}
    Our notion of $\mathcal{M}$-adhesivity follows
    \cite{ehrig2012,ehrig2014adhesive} and is different from the one of
    \cite{azzi2019essence}. What is called $\mathcal{M}$-adhesivity in
    that paper corresponds to our strict
    $\mathcal{M}$-adhesivity. Moreover, in \cite{azzi2019essence} the
    class $\mathcal{M}$ is assumed to be only stable under
    pullbacks. However, if $\mathcal{M}$ contains all split monos, then
    stability under pushouts can be deduced from the other
    axioms~\cite[Prop.~$5.1.21$]{castelnovo2023thesis}.
  \end{remark}
}

@@

We will prove in \Cref{app:ade} two key properties of
$\mathcal{M}$-adhesive categories, $\mathcal{M}$-pushout-pullback
decomposition and uniqueness of pushouts complements, which will be
used in this paper.

\subsection{DPO rewriting systems derivations}
$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using the so-called ``douple pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be a $\mathcal{M}$-adhesive category, a \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $(l,r)$ of arrows with
  the same domain, such that $l$ belongs to $\mathcal{M}$.  The rule
  $\rho$ is said to be \emph{$\mathcal{M}$-linear} if
  $r\in \mathcal{M}$ too. A rule $\rho$ is said to be \emph{consuming}
  if $l$ is not an isomorphism.  The object $L$ is the \emph{left-hand
    side}, $R$ is the \emph{right-hand side} and $K$ the \emph{glueing
    object}.

  A \emph{left-linear DPO rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $R$ is a set of left
  $\mathcal{M}$-linear rules. $(\X, \R)$ will be called \emph{linear}
  if every rule in $R$ is $\mathcal{M}$-linear. Similarly, a
  \emph{consuming} left-linear DPO rewriting system is one in which
  every rule is consuming.

@@

>>>>

This first section is devoted to recall the definition and the basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

@@

We can now define $\mathcal{M}$-adhesive categories.

@@

@@

A number of properties of adhesive categories which play a role in the
theory of rewriting generalise to $\mathcal{M}$-adhesive
categories. These include $\mathcal{M}$-pushout-pullback decomposition
and uniqueness of pushouts complements (details and proofs are in
\Cref{app:ade}).

\subsection{DPO rewriting systems derivations}

$\mathcal{M}$-adhesive categories are the right context in which to
perform abstract rewriting using the so-called ``double pushout
approach'' (DPO). We will recall the basic definitions and properties
of this approach to abstract rewriting.

\begin{definition}[\cite{habel2012mathcal,heindel2009category}]
  Let $\X$ be a $\mathcal{M}$-adhesive category, a \emph{left
    $\mathcal{M}$-linear} rule $\rho$ is a pair $(l,r)$ of arrows with
  $l: K \to L$ and $r: K \to R$, such that $l$ belongs to
  $\mathcal{M}$.  The rule $\rho$ is said to be
  \emph{$\mathcal{M}$-linear} if $r\in \mathcal{M}$ too. A rule $\rho$
  is said to be \emph{consuming} if $l$ is not an isomorphism.  The
  object $L$ is the \emph{left-hand side}, $R$ is the \emph{right-hand
    side} and $K$ the \emph{glueing object}.

  A \emph{left-linear DPO rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $R$ is a set of left
  $\mathcal{M}$-linear rules. The rewiriting system $(\X, \R)$ is
  called \emph{linear} (\emph{consuming}) if every rule in $R$ is
  $\mathcal{M}$-linear (consuming).

@@

Observe that our notion of $\mathcal{M}$-adhesivity follows
\cite{ehrig2012,ehrig2014adhesive} and is different from the one of
\cite{azzi2019essence}. What is called $\mathcal{M}$-adhesivity in
that paper corresponds to our strict
$\mathcal{M}$-adhesivity. Moreover, in \cite{azzi2019essence} the
class $\mathcal{M}$ is assumed to be only stable under
pullbacks. However, if $\mathcal{M}$ contains all split monos, then
stability under pushouts can be deduced from the other
axioms~\cite[Prop.~$5.1.21$]{castelnovo2023thesis}.
